evidence of adaptation-independent time-warped spike timing noise correlations in cortical networks

### 2020-08-08 manuscript.tex
Carry on my wayward son

\title{evidence of adaption-independent time-warped spike timing noise correlations in cortical networks}

@@

\section{Introduction}
The neural coding problem asks how information is encoded through patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}. Whilst precise temporal patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained. Such variability leads to the common assumption of population codes based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
% \cite{gerstner1997neural}, which have capacity beyond the original spike count \cite{}. 
% Continuous firing rates (averaged over time, trial or population ) 
% These population codes assume that trial-to-trial variability is either independent or correlated across the population \cite{averbeck2006neural, lin2015nature}. 
Trial-to-trial spike count variability is in fact correlated across the population in early sensory cortices \cite{reyes2015high, lin2015nature} and was shown to be modulated by a two-dimensional continuous state \cite{lin2015nature}. It is possible that spike time variability is modulated across a population by the same or similar low-dimensional underlying state. This study aims to characterise the nature of trial-to-trial spike time correlations as a step towards confirming or rejecting the modulation of spike times by an underlying low-dimensional latent state. This offers a novel prediction of the encoding capacity of state-dependent spike time encoding. Characterising spike timing codes is important to understanding a variety of dynamic, structural, homeostatic and cognitive processes, as well as properties of action potential patterns directly involved in behaviour.

@@

% Information theoretic approaches aim to understand 
If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a). Neural coding is classically interested in determining the dimensions in which response distributions do not overlap \cite{averbeck2006neural}.

@@

% This paper aims to characterise the precise nature of spike timing noise correlations for single stimuli to uncover the potential encoding capacity of a state-dependent spike time code. 
% The stimulus-dependency of noise correlations is left to future studies.  
Characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}. 
% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
Processes of adaptation, which we define as different to latent states underlying trial-to-trial noise correlations, could also lead to correlative structure in the stimulus-response distribution (Fig. X). Adaptation must therefore be controlled for when characterising adaption-independent noise correlations.

@@

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of spike time noise correlations for first-spike patterns, towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state. The classical `stimulus-dependency' of noise correlations is left to future studies. Previous studies analysing the same dataset found precise multi-neuron intralaminar spike patterns \cite{reyes2014laminar}, spike count noise correlations and could decode stimulus properties (stimulation frequency and locaction) from spike counts and spike patterns \cite{reyes2015high}. The new analysis uses time series analysis to control for neural adaption over all stimulus trials. Adaption occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}, and was confirmed to occur for 7 out of 8 neuron groups in this study.

@@

For repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked. The reader is encouraged to view these plots in the top left of Supplementary Vid. 1. A novel algorithm extracted isolated clusters from these single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}. The novel algorithm, however, also showed that first spike pairs within these isolated clusters were positively correlated above chance-levels.

Moreover, the correlation angle of many of the correlated clusters was different from $45^{\circ}$. This cannot be explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle). Instead the `non-unitary' correlation gradients (different from $45^{\circ}$) may be indicative that correlations are due to an internal shared latent state. Clusters remained correlated with non-unitary gradient after controlling for neural adaption using time series analysis, confirming the correlated clusters as evidence of precise non-adaptive noise correlations. A large proportion of these noise correlations would have been missed without clustering.

Importantly, the observation of adaption-independent correlated clusters with non-unitary gradient shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state. This phenomenon is referred to as `adaption-independent clustered time-warped spike timing'. This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, adaption and different warping functions for different neurons \cite{williams2020discovering}. One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern. Rather, neurons should process relative spike patterns in a state-dependent manner to utilise the improved state-dependent coding capacity.

Adaption-controlled non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers. This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner. Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaption-controlled correlated cluster. This gave a prediction of the resolution of possible state-dependent encoding, which surpasses previously proposed coding forms.

The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. Estimating whether cluster correlations observed for different pairs of neurons is modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came [See notes below]. The work of this paper in extracting and characterising adaption-independent correlated clusters provides a necessary component step towards confirming or rejecting state-dependent spike-time coding, and is also important under the classical fixed-state decoder interpretation. The results show that consideration of clustering, correlative structure and adaption, together, are necessary to explain a large proportion of neural variability.

% It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. 
Testing the state-dependent decoding performance of the clusters is left for future studies, as it would require explicit knowledge of the underlying shared state or an estimate from the spike times of other neurons (which is difficult for the current extracted cluster counts per stimulus in single experimental sessions).
Subthreshold membrane potentials would be a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}. The LFP may offer a useful proxy for the subthreshold membrane potential.

@@

\todo{Look at adaption by neuron type and layer}
\todo{Select final adaption examples, with layers and stats.}
\todo{Do adaption conclusion}
\todo{Make adaption stats overview paragraph simpler, with nice single bar chart}

This section demonstrates that the responses of some neurons exhibit depressive latency adaption, which should be controlled for when testing for correlative structure caused by a non-adaptive state. Figure 2d shows the first spike times of 5 single units in response to repeated stimulations. The L23E unit of Fig. 2d(i) demonstrates an irregular response pattern with adaptation. Whereas the L4I unit of Fig. 2d(ii) shows a short and consistent response, the L… unit of Fig. 2d(iii) reveals a sparse and variable pattern and no obvious adaptation. The L4I and L4I units of Fig. 2d(iv, v) demonstrate more complex response-distribution structure and possible non-linear non-stationarity.

@@

>>>>

\title{evidence of adaptation-independent time-warped spike timing noise correlations in cortical networks}

@@

\section{Introduction}
The neural coding problem addresses the question of how information is encoded by patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}. Whilst precise temporal spike patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained. Such variability leads to the common assumption of population codes based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
% \cite{gerstner1997neural}, which have capacity beyond the original spike count \cite{}. 
% Continuous firing rates (averaged over time, trial or population ) 
% These population codes assume that trial-to-trial variability is either independent or correlated across the population \cite{averbeck2006neural, lin2015nature}. 
Trial-to-trial spike count variability is in fact correlated across the population in early sensory cortices \cite{reyes2015high, lin2015nature} and was shown to be modulated by a two-dimensional continuous state \cite{lin2015nature}. It is possible that spike time variability is modulated across a population by the same or similar low-dimensional underlying state. This study aims to characterise the nature of trial-to-trial spike time correlations as a step towards confirming or rejecting the modulation of spike times by an underlying low-dimensional latent state. This offers a novel prediction of the encoding capacity of state-dependent spike time encoding. Characterising spike timing codes is important to understanding a variety of dynamic, homeostatic and plastic processes, as well as properties of action potential patterns directly involved in cognition.

@@

% Information theoretic approaches aim to understand 
If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a). Neural coding is classically interested in determining the dimensions in which response distributions do not overlap \cite{averbeck2006neural}.

@@

% This paper aims to characterise the precise nature of spike timing noise correlations for single stimuli to uncover the potential encoding capacity of a state-dependent spike time code. 
% The stimulus-dependency of noise correlations is left to future studies.  
Characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}. 
% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
Processes of adaptation, which we define as different to latent states underlying trial-to-trial noise correlations, could also lead to correlative structure in the stimulus-response distribution (Fig. X). Adaptation must therefore be controlled for when characterising adaptation-independent noise correlations.

@@

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of spike time noise correlations for first-spike patterns, towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state. The classical `stimulus-dependency' of noise correlations is left to future studies. A previous study analysing the same dataset found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}. Another study analysing the data set found spike count noise correlations and could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns \cite{reyes2015high}. The new analysis uses time series analysis to control for neural adaptation over all stimulus trials. adaptation occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}, and was confirmed to occur for 7 out of 8 neuron groups in this study.

@@

For repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked (see plots in the top left of Supplementary Vid. 1). A novel algorithm extracted isolated clusters from these single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}. The novel algorithm, however, also showed that first spike pairs within these isolated clusters were positively correlated above chance-levels.

Moreover, the correlation angle of many of the correlated clusters was different from $45^{\circ}$. This cannot be explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle). Instead the `non-unitary' correlation gradients (different from $45^{\circ}$) may be indicative that correlations are due to an internal shared latent state. Clusters remained correlated with non-unitary gradient after controlling for neural adaptation using time series analysis, confirming the correlated clusters as evidence of precise non-adaptive noise correlations. A large proportion of these noise correlations would have been missed without clustering.

Importantly, the observation of adaptation-independent correlated clusters with non-unitary gradient shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state. This phenomenon is referred to as `adaptation-independent clustered time-warped spike timing'. This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, adaptation and different warping functions for different neurons \cite{williams2020discovering}. One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern. Rather, neurons should process relative spike patterns in a state-dependent manner to utilise the improved state-dependent coding capacity.

adaptation-controlled non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers. This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner. Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster. This gave a prediction of the resolution of possible state-dependent encoding, which surpasses previously proposed coding forms.

The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. Estimating whether cluster correlations observed for different pairs of neurons is modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came [See notes below]. The work of this paper in extracting and characterising adaptation-independent correlated clusters provides a necessary component step towards confirming or rejecting state-dependent spike-time coding, and is also important under the classical fixed-state decoder interpretation. The results show that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.

@@

\todo{Look at adaptation by neuron type and layer}
\todo{Select final adaptation examples, with layers and stats.}
\todo{Do adaptation conclusion}
\todo{Make adaptation stats overview paragraph simpler, with nice single bar chart}

This section demonstrates that the responses of some neurons exhibit depressive latency adaptation, which should be controlled for when testing for correlative structure caused by a non-adaptive state. Figure 2d shows the first spike times of 5 single units in response to repeated stimulations. The L23E unit of Fig. 2d(i) demonstrates an irregular response pattern with adaptation. Whereas the L4I unit of Fig. 2d(ii) shows a short and consistent response, the L… unit of Fig. 2d(iii) reveals a sparse and variable pattern and no obvious adaptation. The L4I and L4I units of Fig. 2d(iv, v) demonstrate more complex response-distribution structure and possible non-linear non-stationarity.

@@

\subsection*{Physiological Relevance}
% It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. 
Testing the state-dependent decoding performance of the clusters is left for future studies, as it would require explicit knowledge of the underlying shared state or an estimate from the spike times of other neurons (which is difficult for the current extracted cluster counts per stimulus in single experimental sessions).
Neuronal subthreshold membrane potentials would be a candidate for such state, as they are correlated between neurons across areas \cite{poulet2008internal} and affect action potential initiation \cite{petersen2003interaction}. The LFP may offer a useful proxy for the subthreshold membrane potential.
